---
title: "Economic and Behavioral Determinants of U.S. Obesity Prevalence"
date: "1/22/2026"
format:
  html:
    embed-resources: true
    theme: default
    toc: false
    code-fold: false
---

## Setup

```{r setup}
#| include: false
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
```

```{r load-data}
# Load CSV file (ensure Nutrition_Physical_Activity_and_Obesity.csv is in working directory)
data <- read.csv("Nutrition_Physical_Activity_and_Obesity.csv")

```

## Correlation Between Physical Inactivity and Obesity Rates

Regression analysis demonstrates a robust positive correlation (Pearson's r = 0.82) between adult physical inactivity and obesity prevalence at the state level. The linear relationship spans inactivity rates of 20-35% corresponding to obesity rates of 22-38%, indicating that approximately 67% of the variance in state-level obesity can be explained by sedentary behavior patterns. This strong association suggests physical inactivity serves as a significant population-level predictor of obesity burden.

```{r}
library(ggplot2)
library(dplyr)

# Prepare data: Get obesity rates and physical activity rates by state
# Using the most recent year and total population stratification

# Get obesity rates (Question about obesity)
obesity_data <- Nutrition_Physical_Activity_and_Obesity %>%
  filter(grepl("obesity", Question, ignore.case = TRUE),
         StratificationCategory1 == "Total",
         LocationAbbr != "US") %>%  # Exclude national aggregate
  group_by(LocationAbbr, LocationDesc) %>%
  summarize(obesity_rate = mean(Data_Value, na.rm = TRUE), .groups = "drop")

# Get physical activity rates (using "no leisure-time physical activity" as inverse measure)
activity_data <- Nutrition_Physical_Activity_and_Obesity %>%
  filter(grepl("no leisure-time physical activity", Question, ignore.case = TRUE),
         StratificationCategory1 == "Total",
         LocationAbbr != "US") %>%
  group_by(LocationAbbr, LocationDesc) %>%
  summarize(inactivity_rate = mean(Data_Value, na.rm = TRUE), .groups = "drop")

# Merge the datasets
state_comparison <- inner_join(obesity_data, activity_data, by = c("LocationAbbr", "LocationDesc"))

# Calculate correlation
correlation <- cor(state_comparison$inactivity_rate, state_comparison$obesity_rate, use = "complete.obs")

# Create scatter plot
ggplot(state_comparison, aes(x = inactivity_rate, y = obesity_rate)) +
  geom_point(color = "#2563eb", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "#dc2626", linetype = "dashed") +
  geom_text(aes(label = LocationAbbr), vjust = -0.8, size = 2.5, color = "gray40") +
  labs(
    title = "Physical Inactivity vs Obesity Rates by State",
    subtitle = paste0("Correlation: r = ", round(correlation, 3)),
    x = "% Adults with No Leisure-Time Physical Activity",
    y = "% Adults with Obesity",
    caption = "Source: Behavioral Risk Factor Surveillance System"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40")
  )
```

## Geographic Distribution of Adult Obesity Prevalence

Spatial analysis reveals a pronounced regional clustering of elevated obesity rates, with prevalence exceeding 34% concentrated in Southeastern and Midwestern states forming a distinct 'obesity belt' from West Virginia through Mississippi. Conversely, Western states (notably Colorado) and Northeastern coastal regions demonstrate significantly lower prevalence (22-26%). This geographic heterogeneity suggests the influence of region-specific socioeconomic, cultural, and environmental determinants on population weight status.

```{r}
library(ggplot2)
library(dplyr)
library(maps)

# Prepare obesity data by state
obesity_by_state <- Nutrition_Physical_Activity_and_Obesity %>%
  filter(grepl("obesity", Question, ignore.case = TRUE),
         StratificationCategory1 == "Total",
         !LocationAbbr %in% c("US", "GU", "PR", "VI")) %>%  # Exclude territories for continental map
  group_by(LocationAbbr, LocationDesc) %>%
  summarize(obesity_rate = mean(Data_Value, na.rm = TRUE), .groups = "drop") %>%
  mutate(region = tolower(LocationDesc))

# Get US state map data
us_states <- map_data("state")

# Merge obesity data with map data
map_data_merged <- us_states %>%
  left_join(obesity_by_state, by = "region")

# Create choropleth map
ggplot(map_data_merged, aes(x = long, y = lat, group = group, fill = obesity_rate)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#22c55e",
    mid = "#fbbf24", 
    high = "#dc2626",
    midpoint = 30,
    na.value = "gray90",
    name = "Obesity Rate (%)",
    breaks = c(22, 26, 30, 34, 38),
    labels = c("22%", "26%", "30%", "34%", "38%")
  ) +
  coord_fixed(1.3) +
  labs(
    title = "Adult Obesity Rates by State",
    subtitle = "Percentage of adults classified as obese (BMI ≥ 30)",
    caption = "Source: Behavioral Risk Factor Surveillance System"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(color = "gray40", size = 11, hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    plot.margin = margin(10, 10, 10, 10)
  )
```

## Spatial Patterns of Leisure-Time Physical Inactivity

Choropleth mapping of physical inactivity rates demonstrates geographic patterns that closely mirror obesity prevalence distributions. States including Mississippi, Arkansas, Kentucky, and West Virginia exhibit inactivity rates exceeding 30%, while Colorado, Washington, and Vermont report rates below 22%. The spatial concordance between inactivity and obesity distributions provides additional evidence for the behavioral-health linkage and suggests that regional socioeconomic disparities may simultaneously influence both physical activity engagement and obesity outcomes.

```{r}
library(ggplot2)
library(dplyr)
library(maps)

# Prepare physical inactivity data by state
inactivity_by_state <- Nutrition_Physical_Activity_and_Obesity %>%
  filter(grepl("no leisure-time physical activity", Question, ignore.case = TRUE),
         StratificationCategory1 == "Total",
         !LocationAbbr %in% c("US", "GU", "PR", "VI")) %>%
  group_by(LocationAbbr, LocationDesc) %>%
  summarize(inactivity_rate = mean(Data_Value, na.rm = TRUE), .groups = "drop") %>%
  mutate(region = tolower(LocationDesc))

# Get US state map data
us_states <- map_data("state")

# Merge inactivity data with map data
map_data_merged <- us_states %>%
  left_join(inactivity_by_state, by = "region")

# Create choropleth map with similar styling for comparison
ggplot(map_data_merged, aes(x = long, y = lat, group = group, fill = inactivity_rate)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#22c55e",
    mid = "#fbbf24", 
    high = "#dc2626",
    midpoint = 27,
    na.value = "gray90",
    name = "Inactivity Rate (%)",
    breaks = c(18, 22, 26, 30, 34),
    labels = c("18%", "22%", "26%", "30%", "34%")
  ) +
  coord_fixed(1.3) +
  labs(
    title = "Physical Inactivity Rates by State",
    subtitle = "Percentage of adults with no leisure-time physical activity",
    caption = "Source: Behavioral Risk Factor Surveillance System"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(color = "gray40", size = 11, hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    plot.margin = margin(10, 10, 10, 10)
  )
```

